---
title: "Homework_13"
format: html
---
# Bryan Wilcenski
# 4/15/2026


## Question 1:Does any pair of predictors have combined effects on damage likelihood that are greater than their individual effects? 

# Note: This question and approach was recommended by Professor Burnham after discussing my challenges with developing predictor variables. If there are predictor pairs that act together and have a combined effect on damage probabilities, then it would be possible to use feature engineering to add to my random forest (or Catboost models) to better account for variance. Use of a LLM for coding assistance was also recommended given the number of possible pairs (66). 
```{r}


library(tidyverse)
library(lme4)
library(car)


model_data <- read_csv("Data/model_data_Classifier1.csv")

model_data <- model_data %>%
  mutate(
    Surface = factor(Surface),
    HYDROGROUP = factor(HYDROGROUP),
    PARENT = factor(PARENT),
    StreamCros = factor(StreamCros),
    damaged_bin = ifelse(damaged == "Yes", 1, 0)
  ) #create damage binary (0,1) and assign all categorical variables as factors

head(model_data)


# Base model — all predictors with no interactions
predictors <- c("Avg_Slope", "K_final", "MeanSlope30m", "PercentForestedLand",
                "MEAN", "PM_rank", "Culvert_Count", "Surface", "StreamCros",
                "HYDROGROUP", "PARENT", "Mean_TopoConvergence_50m")

base_formula <- as.formula(paste("damaged_bin ~", paste(predictors, collapse = " + ")))
base_model <- glm(base_formula, data = model_data, family = binomial(link = "logit"))

summary(base_model)
Anova(base_model)

base_aic <- AIC(base_model)

# Screen all 66 pairwise interactions
# Generate all pairs
pairs <- combn(predictors, 2, simplify = FALSE)

# Loop through pairs, store AIC for each interaction model
pair_names <- c()
pair_aics <- c()

for (i in 1:length(pairs)) {
  var1 <- pairs[[i]][1]
  var2 <- pairs[[i]][2]
  int_term <- paste(var1, var2, sep = ":")
  
  int_formula <- as.formula(paste("damaged_bin ~",
                                   paste(predictors, collapse = " + "),
                                   "+", int_term))
  
  int_model <- glm(int_formula, data = model_data, family = binomial(link = "logit"))
  
  pair_names <- c(pair_names, int_term)
  pair_aics <- c(pair_aics, AIC(int_model))
}

# results data frame
results <- data.frame(
  Interaction = pair_names,
  AIC_interaction = round(pair_aics, 2)
)


# create AIC improvement column
results$AIC_improvement <- round(base_aic - results$AIC_interaction, 2)

# Sort best to worst
results <- results[order(-results$AIC_improvement), ]

print(results)

# the best model had a K_final:HYDROGROUP interaction term 
best_model <- glm(damaged_bin ~ Avg_Slope + K_final + MeanSlope30m +
                    PercentForestedLand + MEAN + PM_rank + Culvert_Count +
                    Surface + StreamCros + HYDROGROUP + PARENT +
                    Mean_TopoConvergence_50m +
                    K_final:HYDROGROUP,
                  data = model_data, family = binomial(link = "logit"))

summary(best_model)
Anova(best_model)


# K_final distribution by HYDROGROUP, colored by damage status
ggplot(model_data, aes(x = HYDROGROUP, y = K_final, fill = damaged)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  scale_fill_manual(values = c("No" = "blue", "Yes" = "red"),
                    labels = c("Not Damaged", "Damaged")) +
  labs(title = "Soil Erodibility by Hydrologic Soil Group and Damage Status",
       x = "Hydrologic Soil Group",
       y = "K-Factor (Soil Erodibility)",
       fill = "Status") +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    legend.position = "top"
  )

```

# AIC was used to compare the effects of all interaction terms against the base model with all individual predictors (but no interaction terms) because the large sample size resulted in almost all interactions having a significant p-value. A lower AIC means that the model is better at predicting damages. 

# According to Burnham and Anderson (2004):
# "Some simple rules of thumb are often useful in assessing the relative merits of models in the set: Models having [Delta] i ≤ 2 have substantial support (evidence), those in which 4 ≤ [Delta] i ≤ 7 have considerably less support, and models having [Delta] i > 10 have essentailly no support". 

# In my data, the AIC_Improvement calculation uses the difference between the base model AIC and the interaction term, and thus a higher AIC improvement means that the interaction term improves the model. 

# The greatest improvement was seen when there was a K_final:HYDROGROUP term used in the GLM. While the ANOVA result shows that K_final is not significant on its own (LR Chisq = 0.11), it is significant depending on the HYDROGROUP it is in (LR Chisq = 113.18). While other predictors are powerful on their own, (e.g. MEAN (precipitation), Mean_TopoConvergence_50m (Topographic Convergence), etc.), this finding highlights a new relationship. The boxplot further demonstrates this relationship, showing that on on poorly-drained HYDROGROUP soils (C, C/D) a higher K_factor is associated with more damage. This relationship makes sense--standing water accumulates on the surface, resulting in roads on soils vulnerable to sheet flow (high K_factor) experiencing more damages. 




## Question 2: Does the effect of precipitation on damage probability vary across parent material types?


```{r}



# Mixed model — random intercept and random slope for MEAN by PARENT
mixed_model <- glmer(
  damaged_bin ~ MEAN + MeanSlope30m + Mean_TopoConvergence_50m +
    HYDROGROUP + PercentForestedLand + Avg_Slope + K_final +
    Culvert_Count + StreamCros + Surface +
    (1 + MEAN | PARENT),
  data = model_data,
  family = binomial(link = "logit")
)

summary(mixed_model)
Anova(mixed_model)

re <- as.data.frame(ranef(mixed_model)$PARENT)
re$PARENT <- rownames(re) 
re

barplot(re[,2][order(re[,2])],
        names.arg = re$PARENT[order(re[,2])], #allign orders with the order of the 
        horiz = TRUE, las = 1, #draw horizontal bars and horizontal labels on Y axis
        main = "Precipitation Effect by Parent Material",
        xlab = "Deviation from average effect",
        col = ifelse(re[,2][order(re[,2])] > 0, "firebrick", "steelblue")) #color each bar based on +/-
abline(v = 0, lty = 2)
```

# This is a binomial GLMM with random effects for precipitation by parent material intended to determine if a given amount of precipitation on one soil parent type causes a different probability of damages on another soil type. While some soil parent materials that affect the probability of damage (notably: erosive parent materials like GF and A), the overall changes are small (Variance= 0.007) compared to the the average effect of precipitation (MEAN) across all soils (Estimate = 0.369). This indicates that precipitation's effect on damage probabilities do not differ significantly across parent materials. 